Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-05-27 cordillera-cycle.tex
Improve methods part.

% Copernicus manuscript
%\documentclass[tc, ms]{copernicus}

@@

% Copernicus-like latex2rtf compatible
\input{copernicus_rtf}

@@

% My commands
\newcommand{\idea}[1]{\textbf{[IDEA: #1]}}
\newcommand{\note}[1]{\textbf{[NOTE: #1]}}
\newcommand{\todo}[1]{\textbf{[TODO: #1]}}
\newcommand{\aref}[0]{\textbf{[REF]}}
\def\citep#1{}
\def\citet#1{}

@@

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last glacial
cycle. However, former ice thickness and the ice sheet's contribution to the
last glacial maximum sea-level low-stand remain poorly constrained. Moreover,
our understanding of the glaciation history is mainly restricted to the
deglaciation phase and on regional scale, while little is known about the ice
sheet evolution prior to the last glacial maximum extent. Although
time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
patterns are available for the Laurentide and Eurasian ice sheets, such is not
the case for the Cordilleran ice sheet, where complex arrangements of ice flow
emanating from multiple glaciation centres have been identified.

@@

The simulations presented here were run using Parallel Ice Sheet Model (PISM,
development version~11b0a7f), an open-source, finite-difference, shallow ice
sheet model. The model takes basal topography, sea level, geothermal heat flux
and climate forcing, as input fields and computes the evolution of ice extent
and thickness in time, its thermal and dynamic state, and
the associated lithospheric response. Our modelling domain encompasses the
entire area covered by the Cordilleran ice sheet at the last glacial maximum.
It includes independent ice growing on the western Alaskan ranges and the
Brooks Range and intermediate expanses of ice-free terrain in northern Yukon
Territory and interior Alaska (Fig.~\ref{fig:locmap}).

@@

PISM is a~shallow model, which implies that the balance of stresses is
approximated based on their predominant components.
The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
SIA and SSA velocities are computed by finite difference methods on a~10\,km
resolution horizontal grid of 300 by 150 points (the modelling domain). Ice
softness depends on temperature and water content through an enthalpy
formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is
computed in three dimensions in up to 51 irregularly spaced layers within the
ice, and temperature is additionally computed in 31 regularly spaced layers in
bedrock to a~depth of 3\,km. Surface air temperature from the climate forcing
provides the upper boundary condition to the ice enthalpy model, and
a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary
condition to the bedrock thermal model. Although this uniform value does not
account for the high spatial geothermal variability in the region, it is on
average representative of available heat flow measurements
\citep{artemieva-mooney-2001,blackwell-richards-2004}. Whereas spatial
variations in the geothermal heat flux may significantly affect basal
conditions \citep{pattyn-2010}, this study focuses on reconstructing the former
extent of the Cordilleran ice sheet, on which it has relatively little effect
\citep{rogozhina-etal-2012}.

\todo{Describe the new sliding law?}

A~pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the
bed-parallel shear stress and the sliding velocity. The yield stress is
modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
varies from 15 to 45{\degree}. It is computed once at the beginning of the run,
as a~function of modern bed elevation, with lowest values occurring at low
elevations,

\begin{align}
  \phi =
  \begin{cases}
    15      & \text{for }       z<  0 \\
    z/15+15 & \text{for }    0 <z<200 \\
    45      & \text{for }  200 <z     \\
  \end{cases}
\end{align}

where $\phi$ is given in degrees and $z$ in meters above current sea level.
This model thus accounts for a~weakening of till associated with the presence
of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}. Because the
distribution of marine sediments is assumed constant throughout the run, the
till friction angle $\phi$ is not affected by changes in bed elevation. Basal
topography is derived from the ETOPO1 combined topography
and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.

@@

Ice surface accumulation and ablation are computed from monthly mean surface
air temperature and monthly precipitation by a~temperature-index (positive
degree-day) model \citep{hock-2003}. Ice accumulation is equal to precipitation
when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly
with temperature between 0 and 2\,\unit{{\degree}C}. Ice ablation is computed
from the number of positive degree-days, defined as the integral of
temperatures above 0\,\unit{{\degree}C} in one year.

The positive degree-day integral \citep{calov-greve-2005} is numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is part of the forcing climatology and was
computed from daily temperature values from the North American Regional
Reanalysis \citep{data:narr} and excludes variability associated with the annual
temperature cycle itself \note{improvement} \citet{seguinot-rogozhina-2014}. The
ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{shea-etal-2009}.

@@

Atmospheric forcing of the model consists of a present-day monthly climatology
modified by a lapse-rate correction and time-dependent temperature offsets. The
present-day climatology was computed from the near-surface air temperature and
precipitation rate fields of the North American Regional Reanalysis (NARR) over
the period xxxx--xxxx (Fig.~\ref{fig:atm}). Modern climate of the North
American Cordillera is characterised by strong geographic variations of
seasonality, timing of the precipitation peak and daily temperature variability.
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as outlined in a previous sensitivity
study.

Temperature offset time-series are derived from proxy records from
the Greenland Ice Core Project (GRIP), the North Greenland Ice Core Project
(NGRIP), the European Project for Ice Coring in Antarctica (EPICA), the Lake
Vostok ice core, and Ocean Drilling Program (ODP) sites 1012 and 1020, both
located off the Californian shore. Palaeo-temperatures from the GRIP and NGRIP
records were calculated using a quadratic equation \aref:

\todo{add equation here,}

while the temperature reconstructions from the Antarctic and oceanic cores were
provided by their authors. All records were scaled linearly in
order to simulate comparable ice volumes and realistic ice extents at the last
glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).

>>>>

% Copernicus manuscript
\documentclass[tc, ms]{copernicus}

@@

% Copernicus-like latex2rtf compatible
%\input{copernicus_rtf}

@@

% My commands
\newcommand{\idea}[1]{\textbf{[IDEA: #1]}}
\newcommand{\note}[1]{\textbf{[NOTE: #1]}}
\newcommand{\todo}[1]{\textbf{[TODO: #1]}}
\newcommand{\aref}[0]{\textbf{[ref.]}}
\renewcommand{\citep}[1]{\aref}
\renewcommand{\citet}[1]{\aref}

@@

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last glacial
cycle. However, former ice thickness and the ice sheet's contribution to the
last glacial maximum sea-level low-stand remain poorly constrained. Moreover,
our understanding of the glaciation history is mainly restricted to the
deglaciation phase and on a regional scale, while little is known about the ice
sheet evolution prior to the last glacial maximum extent. Although
time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
patterns are available for the Laurentide and Eurasian ice sheets, such is not
the case for the Cordilleran ice sheet, where complex arrangements of ice flow
directions emanating from multiple glaciation centres have been identified.

@@

The simulations presented here were run using Parallel Ice Sheet Model (PISM,
development version~11b0a7f), an open-source, finite-difference, shallow ice
sheet model. The model takes basal topography, sea level, geothermal heat flux
and climate forcing, as input fields and computes the evolution of ice extent
and thickness in time, its thermal and dynamic state, and
the associated lithospheric response. Our modelling domain encompasses the
entire area covered by the Cordilleran ice sheet at the last glacial maximum
(Fig.~\ref{fig:locmap}).

@@

PISM is a~shallow model, which implies that the balance of stresses is
approximated based on their predominant components.
The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
SIA and SSA velocities are computed by finite difference methods on a~10\,km
resolution horizontal grid of 300 by 150 points (the modelling domain). Ice
softness depends on temperature and water content through an enthalpy
formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is
computed in three dimensions in up to 51 layers within the
ice, and temperature is additionally computed in 31 layers in bedrock to
a~depth of 3\,km. Surface air temperature from the atmospheric forcing
provides the upper boundary condition to the ice enthalpy model, and
a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary
condition to the bedrock thermal model. Although this uniform value does not
account for the high spatial geothermal variability in the region, it is on
average representative of available heat flow measurements
\citep{artemieva-mooney-2001,blackwell-richards-2004}.

A~pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the
bed-parallel shear stress and the sliding velocity. The yield stress is
modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
varies from 15 to 45{\degree}. It is taken as a~function of modern bed
elevation, with lowest values occurring at low elevation, thus accounting
for a~weakening of till associated with the presence of marine sediments
\citep{martin-etal-2011,aschwanden-etal-2013}. The till effective pressure is
related to the modelled amount of water in the till, using a formula derived
from laboratory experiments with till extracted from an Antarctic ice stream.
Basal topography is derived from the ETOPO1 combined topography
and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.

\note{Do we need more details (equations) on this new sliding law?}

@@

Ice surface accumulation and ablation are computed from monthly mean
near-surface air temperature, monthly precipitation, and monthly standard
deviation of near-surface temperature by a~temperature-index (positive
degree-day) model \citep{hock-2003}. Ice accumulation is equal to precipitation
when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly
with temperature between 0 and 2\,\unit{{\degree}C}. Ice ablation is computed
from the number of positive degree-days, defined as the integral of
temperatures above 0\,\unit{{\degree}C} in one year.

The positive degree-day integral \citep{calov-greve-2005} is numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is part of the forcing climatology. It was
computed from daily temperature values from the North American Regional
Reanalysis \citep{data:narr}, excluding variability associated with the
seasonal cycle itself \citet{seguinot-rogozhina-2014}. The
ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{shea-etal-2009}.

@@

Atmospheric forcing of the model consists of a present-day monthly climatology
modified by a lapse-rate correction and time-dependent temperature offsets. The
present-day climatology was computed from the near-surface air temperature and
precipitation rate fields of the North American Regional Reanalysis (NARR) over
the period 1979--2000 (Fig.~\ref{fig:atm}). Modern climate of the North
American Cordillera is characterised by strong geographic variations of
seasonality, timing of the precipitation peak and daily temperature variability.
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as outlined in a previous sensitivity
study.

Temperature offset time-series are derived from proxy records from
the Greenland Ice Core Project (GRIP), the North Greenland Ice Core Project
(NGRIP), the European Project for Ice Coring in Antarctica (EPICA), the Lake
Vostok ice core, and Ocean Drilling Program (ODP) sites 1012 and 1020, both
located off the Californian shore. Palaeo-temperatures from the GRIP and NGRIP
records were calculated using a quadratic equation \aref:
\begin{equation}
    \Delta T = -11.88(\chem{\delta^{18}O}-\chem{\delta^{18}O}_0)
               -0.1925(\chem{\delta^{18}O}^2-\chem{\delta^{18}O}_0^2)
\end{equation}
while the temperature reconstructions from the Antarctic and oceanic cores were
provided by their authors. All records were scaled linearly in
order to simulate comparable ice volumes and realistic ice extents at the last
glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).
